---
title: "vl_final"
output: html_document
---

# Comparison of peptide-MHC-I Binding Prediction Models

Author: Vishal Lashkari

Purpose: to use descriptive and inferential statistics to compare the predictions made by well-established models (NetMHCpan) against predictions made by models with newer architectures (HLApollo).

# Tile Protein Sequences

From protein sequences in fasta format, create a tiled csv of peptides suitable for predictions on NetMHCpan and HLApollo.

Here, human chromosome 21 is tiled into 9mers. Duplicates are removed. Some duplicate peptides were retained if the flanking regions are distinct as this is an additional input to HLApollo.

No pathogenic peptides are included.

Data from UniProt: <https://www.uniprot.org/proteomes/UP000005640>

```{r}
library(readr)
library(dplyr)
library(Biostrings)

proteome<-readAAStringSet('chr21.fasta')

flank_len <- 10   # length of flanks (10 for HLApollo)
mer <- 9          # peptide length
pred_allele <- 'A*02:01' # To notate for HLApollo

# Empty list to store tiles
results <- list()

# Hash to prevent duplicates
seen <- new.env(hash = TRUE, parent = emptyenv())

# Create a progress bar
bar <- txtProgressBar(min = 0, max = length(proteome), style = 3)

# For each protein sequence in the fasta
for (i in 1:length(proteome)) {
  seq <- proteome[i]  # get the protein sequence
  prot_len <- width(seq)
  
  # Unlikely to exist but skip proteins with length < peptide length
  if (prot_len < mer) {
    setTxtProgressBar(bar, i)
    next
  }
  
  # Sliding window
  for (start_pos in 1:(prot_len - mer + 1)) {
    query <- subseq(seq, start_pos, start_pos + mer - 1) # seq to be predicted
    
    # store N terminal flanking residues
    n_start <- max(1, start_pos - flank_len)
    n_flank <- subseq(seq, n_start, start_pos - 1)
    n_flank <- as.character(n_flank)
    
    # Pad with '*' for HLApollo
    if (nchar(n_flank) < flank_len){
      n_flank <- paste0(strrep('*', flank_len - nchar(n_flank)), n_flank)
    }
    
    # store C terminal flanking residues
    c_end <- min(prot_len, start_pos + mer + flank_len - 1)
    c_flank <- subseq(seq, start_pos + mer, c_end)
    c_flank <- as.character(c_flank)
    
    # Pad with '*' for HLApollo
    if (nchar(c_flank) < flank_len){
      c_flank <- paste0(c_flank, strrep('*', flank_len - nchar(c_flank)))
    }
    
    # Hash the peptide, n flank, and c flank
    key <- paste(n_flank, query, c_flank, sep = '|')
    
    # Do not store duplicate triplets (may have duplicate sequences)
    if (exists(key, envir = seen, inherits = FALSE))
      next
    
    assign(key, TRUE, envir = seen)
    
    # Store
    results[[length(results) + 1]] <- list(
      allele = (pred_allele),
      peptide = as.character(query),
      n_flank = as.character(n_flank),
      c_flank = as.character(c_flank)
    )
  }
  
  # Update progress bar for each protein
  setTxtProgressBar(bar, i)
}

# Close the progress bar
close(bar)

# Aggregate into df
proteome_flanks <- bind_rows(results)

# Write csv (in HLApollo input format)
write_csv(proteome_flanks, "chr21_tiled.csv")

```

# Make Predictions

Predictions on the tiled chromosome were made with:

-   NetMHCpan 4.1 <https://services.healthtech.dtu.dk/services/NetMHCpan-4.1/>
-   HLApollo <https://github.com/Genentech/HLApollo>

Predictions were made only for HLA-A\*02:01, the most common allele in the United States.

The relevant metric is the percentile rank which both models output. A stronger binder will have a lower value for its percentile rank.

NetMHCpan predictions were made only on the 9mer sequence in question.

HLApollo predictions were made by additionally supplying the 10 aa flanking the 9mer on either side in the native protein sequence.

# Combine Predictions

Percentile ranks for the same 9mer but with different flanking residues were averaged. This produces a list of unique peptides, each with two percentile predictions (one from each model).

```{r}
library(readr)
library(tidyverse)

# Load predictions data
net <- read_csv('chr21_tiled_netmhcpan.csv')
apollo <- read_csv('chr21_tiled_hlapollo.csv')
nrow(net)
nrow(apollo)
# Remove duplicates from net (predicts on peptide only without flanks)
net <- net %>%
  distinct(Peptide, .keep_all = TRUE)

# Average duplicate from apollo (peptides with different flanks)
apollo <- apollo %>%
  group_by(peptide) %>%
  summarise(mhc_pred_0_rank = mean(mhc_pred_0_rank), .groups = 'drop')

nrow(apollo)

# Merge the predictions
# SELECT net.Peptide AS peptide
# net.EL_rank AS net,
# apollo.mhc_pred_0_rank AS apollo
# FROM net
# JOIN apollo
# ON net.Peptide = apollo.peptide

predictions <- net %>%
  inner_join(apollo, by=c('Peptide'='peptide')) %>%
  transmute(peptide = Peptide, net = EL_Rank, apollo = mhc_pred_0_rank)

# Should match length of net because net has no duplicate peptides
# apollo has some duplicate peptides if they have unique flanks
# After merging, apollo length should also match
nrow(predictions)

head(predictions)
```

# Density Plot

Show the distribution of predictions for each model on the same set of peptides.

Also, show the median percentile rank.

```{r}
library(ggplot2)
library(tidyr)

# Create density plot

pred_long <- predictions %>%
  pivot_longer(cols = c(net, apollo), names_to = "model", values_to = "score") %>%
  mutate(model = recode(model, net = "NetMHCpan", apollo = "HLApollo"))

# Calculate median to annotate
medians <- pred_long %>%
  group_by(model) %>%
  summarise(median = median(score))

den <- ggplot(pred_long, aes(x = score, fill = model)) +
  geom_density(alpha = 0.4) +
  geom_vline(data = medians, aes(xintercept = median, color = model),
             linetype = "dashed", size = 1, show.legend=FALSE) +
  geom_text(data = medians, aes(x = median, y = 0.07, label = paste0("Median: ", round(median,2))),
            angle = 90, vjust = -0.5, hjust = 1, color = "black") +
  labs(title = "Distribution of Predicted Binding", x = "Binder Percentile", 
       y = "Density", fill = "Model")

colors <- ggplot_build(den)$data[[1]]$fill %>% unique()

print(den)
```

# Comparison of Strong Binders

Call strong binders as those predicted in the top 1%.

```{r}
# call strong binders as top 1%
venn_predictions <- predictions %>%
  mutate(
    net_strong = net < 1,
    apollo_strong = apollo < 1
  )

net_strong_count <- sum(venn_predictions$net_strong)
apollo_strong_count <- sum(venn_predictions$apollo_strong)
overlap_count <- sum(venn_predictions$net_strong & venn_predictions$apollo_strong)

library(ggvenn)

venn_data <- list(
  NetMHCpan = venn_predictions$peptide[venn_predictions$net_strong],
  HLApollo  = venn_predictions$peptide[venn_predictions$apollo_strong]
)

ggvenn(venn_data,
       fill_color = colors) +  # match histogram colors
  ggtitle("Predicted Strong Binders")


```

# Model Agreement

Investigate model agreement with a Bland-Altman plot since two different models are predicting the same metric on the same peptides.

```{r}
library(blandr)

# Bland–Altman plot
diffs <- predictions$net - predictions$apollo
mean_diff <- mean(diffs); sd_diff <- sd(diffs)
blandr.draw(predictions$net, predictions$apollo) +
  labs(title="Model Agreement", x="Mean", y="Difference") +
  geom_hline(yintercept = mean_diff, linetype="solid") +
  geom_hline(yintercept = mean_diff + 1.96*sd_diff, linetype="dashed") +
  geom_hline(yintercept = mean_diff - 1.96*sd_diff, linetype="dashed") +
  annotate("text", x=11, y=c(mean_diff, mean_diff + 1.96*sd_diff, mean_diff - 1.96*sd_diff),
           label = paste0(c("Mean: ", "Upper: ", "Lower: "), round(c(mean_diff, mean_diff + 1.96*sd_diff, mean_diff - 1.96*sd_diff),2)),
           hjust=1, vjust=c(-0.5,-0.5,1.5))


```

# Model Correlation

The models may have correlation without agreement.

```{r}
library(ggplot2)

cor_val <- cor(predictions$net, predictions$apollo)

ggplot(predictions, aes(x = net, y = apollo)) + 
  geom_point(alpha = 0.2) + 
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  annotate("text", x = 100, y = 100,
           label = paste0("r = ", round(cor_val, 2)), color = "black") +
    labs(title = "Comparison of Predicted Binder Percentiles",
       x = "NetMHCpan Percentile",
       y = "HLApollo Percentile")
```

# Hypothesis Testing

Use a paired t-test as well as the Wilcoxon signed-rank test to determine if the two models have statistically significantly different predictions.

```{r}
# Paired t-test
print(t.test(predictions$net, predictions$apollo, paired = TRUE))

# Wilcoxon
print(wilcox.test(predictions$net, predictions$apollo, paired = TRUE))

```
